Density-dependent selection with strictly relative contests

### 2017-11-11 manuscript.tex
latest
l

\title{A lottery model of density-dependent selection}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}.

Despite this variety, fitness measures are different ways to environmentally-dependent quantify the differential reproduction and survival of individuals \citep{metcalf_2007}.

relative fitness is almost entirely divorced from population ecology.

In uncrowded populations, there is no discrepancy. Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp. 26]{crow_1970}.

Absolute measures of fitness like $r$ and $K$ are derived directly from population ecology. By contrast,

This discrepancy runs deeper than a simple difference in emphasis on absolute densities rather than relative frequencies.

 The discrepancy emerges in crowded populations. Then the relative fitness description of selection assumes that $N$ is fixed, and fitness strictly involves the ability to increase in frequency at the expense of others (e.g. in the Wright-Fisher model). On the other hand, from quite general population ecological considerations, MacArthur famously argued that ``fitness is $K$'' in crowded populations ($K$-selection) \citep[pp. 149]{macarthur_1967}. To appreciate the import of MacArthur's argument, and see the nature of the discrepancy between these views of selection, we now briefly review MacArthur's argument.

MacArthur considers two types subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. For a type to have the possibility of persisting, it must be able to grow in some region of $(n_1,n_2)$ space at higher total population density $N=n_1+n_2$ than the other type (Fig.~\ref{fig:Ksel}a).

Clearly MacArthur's argument is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b). Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$

In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist.

Given the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}, at this point we might start to doubt that the relative fitness, constant=$N$ description of selection has a plausible basis in population ecology. In particular, the uncoupling of selection and population demography could be problematic, because they are both driven by the same birth/death events.

In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all. The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection. But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines. For instance, contests for territory and mates are predominantly relative. MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption. The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.

Here we investigate the

territorial model of growth, dispersal and competition.

@@

Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two genotypes). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}
To supplement our analytical mean field derivation, we did numerical simulations of our exact our density-dependent lottery model, and verified  that Eq. \eqref{eq:master} is a good approximation (Appendix B). Thus, Eq. \eqref{eq:master} describes how type abundances change over time in a lottery model where population density can itself vary with time.

@@

>>>>

\title{Density-dependent selection with strictly relative contests}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. This variety is not problematic by itself because different fitness measures serve different purposes. But ultimately any measure of fitness should be grounded in the processes of birth and death which govern population ecology and demography \citep{metcalf_2007}. While this connection is apparent for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.

In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}, with selection favoring greater $r$ ($r$-selection). In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course. This is exemplified by the Wright-Fisher model, in which time advances in discrete generations which do not overlap, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004}). But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large. The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. For a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In this case the $f_1$ and $f_2$ nullclines are not even defined. Moreover, at face value MacArthur's, conclusion that ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} seems to be incompatible with constant $N$. In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in $N$. More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density. This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and the lab.

Here we introduce a novel density-dependent model of population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. We show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations when MacArthur's very liberal use of ``$K$'' is taken into account. In particular, MacArthur's argument does not justify the widespread intuition that selection in crowded environments  attempts to maximize population size is a consequence of

argument is widely interpreted to mean that positive selection in crowded populations involves the ability to keep growing at ever higher population densities [citations]

Moreover, it is not compatible with most verbal interpretations of MacArthur's argument,

\citep{anderson_1971}

 with the assumption of constant $N$, although as we will discuss below, it is a narrow interpretation of MacArthur's argument.

@@

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two genotypes). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}.\label{eq:Da}
\end{equation}
To supplement our analytical mean field derivation, we did numerical simulations of our exact our density-dependent lottery model, and verified  that Eq. \eqref{eq:master} is a good approximation (Appendix B). Thus, Eq. \eqref{eq:master} describes how type abundances change over time in a lottery model where population density can itself vary with time.

@@

%In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all. The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection. But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines. For instance, contests for territory and mates are predominantly relative. MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption. The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.
